A Constraint Solving Approach to Parikh Images of Regular Languages

### 2022-07-08 main.tex
intro

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{numprint}
\usepackage{ebproof}

@@

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in the tool \Catra. \Catra also supports the approximate method of
\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata intersection problems generated by
\OstrichPlus{} when solving the PyEx string constraint benchmark suite involving
string length constraints \cite{pyex}.

@@

Parikh image computations as well as Parikh automata feature extensively in
string solvers, including as mentioned above \Ostrich{} and \OstrichPlus{}
\cite{ostrich,ostrich-plus}, but also formes the basis of Trau~\cite{trau-pldi},
and occurs in \textsc{Sloth}~\cite{sloth}. Parikh images frequently appear when
introducing cardinality constraints like length or string indexing. The
state-of-the-art approach to handling Parikh image computation is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Automaton_1 \times \ldots \times \Automaton_k)$ with the
conjunction of the automata's parikh maps, $\AndComp{i=1}^{k}
\ParikhMap(\Automaton_i)$. This approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}.

@@

Many other generalisations of the Parikh than the ones we use here have been
studied. Prominent examples include generalising the Parikh map to segments of a
fixed length \cite{KARHUMAKI1980155} and the more general Parikh matrix, which
gives more information about a word than the standard Parikh image
\cite{parikh-matrix}. Another notable generalisation is the p-vector, introduced
in~\cite{infinite-words}, which denotes the position of each letter in the word
rather than the number of their occurrences and allows for generalisations into
infinite alphabets. All of these in some sense extend the Parikh map. By
contrast, the main utility of the formulation introduced here is that it allows
us to \emph{remove} something, thereby potentialy obtaining an easier problem.

@@

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{/}\SndRegex{}\mathtt{/} \text{ accepts } s\\
  & \mathtt{/}\FstRegex{}\mathtt{/} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

@@

@@

\subsection{Producing the Parikh image}\label{sec:introduction:parikh}

>>>>

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage[capitalise]{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{numprint}
\usepackage{ebproof}

@@

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in the tool \Catra. \Catra{} also supports the approximate method of
\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata intersection problems generated by
\OstrichPlus{} when solving the PyEx string constraint benchmark suite involving
string length constraints \cite{pyex}.

@@

Parikh image computations as well as Parikh automata feature extensively in
string solvers, including as mentioned above \Ostrich{} and \OstrichPlus{}
\cite{ostrich,ostrich-plus}, but also forms the basis of Trau~\cite{trau-pldi},
and occurs in \textsc{Sloth}~\cite{sloth}. Parikh images frequently appear when
introducing cardinality constraints like length or string indexing. The
state-of-the-art approach to handling Parikh image computation is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Language(\Automaton_1) \cap \ldots \cap \Language(\Automaton_k))$ with the
conjunction of the automata's parikh maps, $\bigwedge_{i=1}^{k}
\ParikhMap(\Automaton_i)$. This approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}.

@@

Many other generalisations of the Parikh image than the projections we use here have been
studied. Prominent examples include generalising the Parikh map to segments of a
fixed length \cite{KARHUMAKI1980155} and the more general Parikh matrix, which
gives more information about a word than the standard Parikh image
\cite{parikh-matrix}. Another notable generalisation is the p-vector, introduced
in~\cite{infinite-words}, which denotes the position of each letter in the word
rather than the number of their occurrences and allows for generalisations into
infinite alphabets. All of these in some sense extend the Parikh map. By
contrast, the main utility of the formulation introduced here is that it allows
us to \emph{remove} something, thereby potentialy obtaining an easier problem.

@@

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\[
   s \in \Language(\mathtt{/}\SndRegex{}\mathtt{/})  \land
   s \in \Language(\mathtt{/}\FstRegex{}\mathtt{/})  \land
   \MonoidElement = f(s)  \land
   \varphi(\MonoidElement)
\]

@@

  \caption{The collection of automata we use as running examples in this paper, with annotations illustrating intermittent states of our calculus.}    

@@

\subsection{Satisfying Constraints on Occurrence of Letters}\label{sec:introduction:parikh}
